Final draft

### 2021-07-12 final draft.Rmd

---

\chapter*{Introduction}

@@

>>>>

---
```{r, results= FALSE, message=FALSE, warning=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.pos = '')
knitr::opts_chunk$set(out.extra='')
```
```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
Sys.setlocale("LC_TIME", "English")
```
```{r include=FALSE}
library(ggplot2)
library(ggpubr)
library(kableExtra)
```
\chapter*{Introduction}

@@

\chapter{Applications to Stochastic Volatility Models}
\section{Comparing Constant and Stochastic Volatility Models of Equity Returns}

In this section, we apply the particle filter in a model-comparison
exercise, ultimately underlining the importance of taking stochastic
volatility into account in modeling equity returns.

We consider a stochastic volatility model
$$
\mathcal{M}_{SV}:\begin{array}{lc}
r_{t}=\alpha_{r}+\sigma_{r,t}\cdot v_{t}\\
\log\sigma_{r,t}^{2}=\alpha_{\sigma}+\beta_{\sigma}\cdot\log\sigma_{r,t-1}^{2}+\sigma_{\sigma}\cdot w_{t}
\end{array}\ (v_{t},w_{t})\sim\mathcal{N}_{2}(\boldsymbol{0},I_{2})
$$

and a constant volatility iid model 
$$
\mathcal{M}_{CV}:(r_{t})\overset{iid}{\sim}\mathcal{N}(\alpha,\sigma^{2}).
$$

We calibrate the model parameters using the whole sample, and proceed
by setting the return mean of both models to equal the sample mean,
$\sigma^{2}$ equal to the return variance, and the parameters of
the volatility equation in $\mathcal{M}_{SV}$ by running an $AR(1)$
regression of the log squared residuals.

In order to compare the predictive performance of the two, we compare
the cumulative log likelihood of the observed returns up to time $t$
given the considered model. In formulas, the cumulative likelihood
can be computed as
$$
p(y_{1:t}|\mathcal{M}_{i})=\prod_{s=0}^{t-1}p(y_{s+1}|y_{1:s},\mathcal{M}_{i}),i\in\{SV,CV\},
$$

adopting the convention $y_{1:0}=\emptyset$. This is trivial to obtain
in the iid case; considering the stohastic volatility model, the predictive
likelihoods can be derived as

$$
p(y_{t+1}|y_{1:t},\mathcal{M}_{SV})=\int p(y_{t+1}|\sigma_{r,t+1},y_{1:t},\mathcal{M}_{SV})p(\sigma_{r,t+1}|y_{1:t},\mathcal{M}_{SV})d(\sigma_{r,t+1})
$$
$$
=\int p(y_{t+1}|\sigma_{r,t+1}\mathcal{M}_{SV})p(\sigma_{r,t+1}|y_{1:t},\mathcal{M}_{SV})d(\sigma_{r,t+1})
$$

Whereas the emission distribution $p(y_{t+1}|\sigma_{r,t+1}\mathcal{M}_{SV})$
is known, we approximate the predictive state distribution $p(\sigma_{r,t+1}|y_{1:t},\mathcal{M}_{SV})$
with weighted particles obtained with particle filtering.

Figure XX reports the time series of the cumulative log-likelihood
given the constant volatility model $\mathcal{M}_{CV}$ relative to
the one given $\mathcal{M}_{SV}$, obtained as $\log p(y_{1:t}|\mathcal{M}_{CV})-\log p(y_{1:t}|\mathcal{M}_{SV})$.
Notice that the first difference of this series is the relative log-likelihood
of the incoming observation, $p(y_{t}|y_{1:t-1},\mathcal{M}_{CV})-p(y_{t}|y_{1:t-1},\mathcal{M}_{SV})$.

```{r cll_plot, echo=F, fig.align='center', fig.cap="Relative log-likelihood", fig.height=3, fig.pos='H', fig.width=10, results=F}
load("SV/cll_plot.Rda")
cll_plot+
  theme_bw()+
  scale_x_date(date_breaks = "3 months",date_minor_breaks="1 month",date_labels = "%m/%y")+
  labs(x="Time (month/year)",y="")
```
Overall, this analysis shows how a model with stochastic
volatility is better at describing and predicting returns. As can be seen from the graph and perhaps unsurprisingly,
the density forecast given the stochastic volatility model represents
a particularly significant improvement in the period of extreme values of returns in March 2020.
